\vspace{-2.2cm}Multilevel selection in causal models: the multiplicative nature of evolutionary and probabilistic selection processes as the general driver for the emergence of cooperation and specialization (Summary paper).\footnote{Add stable link to this summary file}

### 2022-02-14 summary.tex
Summary

%opening
\title{\vspace{-2.2cm}Multilevel selection in causal models: the non-ergodicity of evolutionary and probability theories as the general driver for the emergence of cooperation and specialization.\footnote{\url{https://github.com/glandfried/transitions/releases/download/doc.0.0.0/landfried-trans\en{itions}\es{iciones}.pdf} } }
\author{Gustavo Landfried$^1$}
\affil{\small 1. Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Computaci\'on. Buenos Aires, Argentina}
\affil{\en{Mail}\es{Correo:} \url{glandfried@dc.uba.ar}}

@@

\begin{abstract}
\en{In the last third of the universe's history, a simple self-replicating organization of matter emerged on earth. }%
\es{En el último tercio de la historia del universo surgió en la tierra una organización de la materia simple capaz de autoreplicarse. }%
%
\en{The growth of these lineages followed multiplicative and noisy processes: sequences of survival and reproduction rates. }%
\es{El crecimiento de estos linajes siguieron procesos multiplicativos y ruidosos: secuencias de tasas de supervivencia y reproducción. }%
%
% \en{The errors produced during replication diversified the life forms, and the growth rates of the different strategies favored those better adapted to the environment. }%
% \es{Los errores producidos durante la replicación diversificaron las formas de vida, y las tasas de crecimiento de las diferentes estrategias favorecieron a aquellas mejor adaptadas al ambiente. }%
% %
\en{The current complexity of life is the consequence of a series of evolutionary transitions in which entities capable of self-replication after the transition become part of higher level cooperative units. }%
\es{La complejidad actual de la vida es consecuencia de una serie de transiciones evolutivas en las que entidades capaces de autoreplicación luego de la transición pasan a formar parte de unidades cooperativas de nivel superior. }%
%
% \en{How to explain this permanent tendency of life in favor of cooperative aggregation and specialization? }%
% \es{¿Cómo se explica esta tendencia permanente de la vida en favor de la agregación cooperativa y la especialización? }%
\en{To explain evolutionary transitions, it is necessary to demonstrate the evolutionary advantage of cooperation, but also the advantage of specialization. }%
\es{Para explicar las transiciones evolutivas es necesario demostrar la ventaja evolutiva de la cooperación, pero también la ventaja de la especiliazación. }%

@@

\en{Recently it has been shown that, as a consequence of the non-ergodicity of multiplicative processes, fluctuations have a negative effect on individual growth rates, which can be reduced through mutual cooperation, leading to an increase in the long-term growth rates of all individuals. }%
\es{Recientemente se ha mostrado que, como consecuencia de la no-ergodicidad de los procesos multiplicativos, las fluctuaciones tienen un efecto negativo en las tasas de crecimiento individuales, las cuales pueden ser reducidas a través de la mutua cooperación, produciendo un aumento de la tasas de crecimiento a largo plazo de todos los individuos. }%
%
% \en{Ole Peters believs that the increase in the growth rate is sufficient argument to demonstrate the evolutionary advantage of cooperation, which he proposes as the main explanation for evolutionary transitions. }%
% \es{Ole Peters considera que el aumento de la tasa de crecimiento es argumento suficiente para demostrar la ventaja evolutiva de la cooperación, lo que propone como principal explicación de las transiciones evolutivas. }%
% %
% \en{However, he does not consider the problem of defection, who says ``our cooperators are unable to break the cooperative pact''. }%
% \es{Sin embargo no considera el problema de la deserción, quien dice ``our cooperators are unable to break the cooperative pact''. }%
% %
% \en{Yaari suggests a group selection argument but he did not provide a formal demonstration. }
% \es{Yaari sugiere un argumento de selección de grupos pero no prové una demostración formal. }
\en{However, these papers do not provide a formal proof of the evolutionary advantage of cooperation. }%
\es{Sin embargo estos trabajos no porveen una demostración formal que justifique la ventaja evolutiva de la cooperación. }%
% Parrafo
%
\en{For this purpose it is necessary to consider selection at both the individual and group level. }%
\es{Para este propósito es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{The co-author of the concept of evolutionary transitions (Szathmáry) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models, making use of the isomorphism between evolutionary theory and Bayesian inference. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmary) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%

@@

\en{In this paper we specify a probabilistic causal model, in which individuals are affected by the environment and by the social behaviors of cooperation and defection of their context. }%
\es{En este trabajo especificamos un modelo causal probabilístico, en el que los individuos se ven afectados por el ambiente y por los comportamientos sociales de cooperación y deserción de su contexto. }%
\en{Under this minimal set of hypotheses, where we consider \emph{unconditionally} cooperative individuals who generate a common good that can be exploited by defecting individuals without receiving some kind of punishment in return (e.g. end of cooperation), the evolution of cooperation literature predicts defection as the only evolutionarily stable strategy. }%
\es{Bajo este conjunto mínimo de hipótesis, donde consideramos individuos \emph{incondicionalmente} cooperadores que generan un bien común que puede ser explotado por individuos desertores sin que reciban a cambio algún tipo de castigo (e.g fin de la cooperación), la literatura de evolución de la cooperación predice que la deserción es la única estrategia evolutivamente estable. }%
% \en{It is assumed that common goods, if not accompanied by special conditions (such as communication allowing coordination, memory allowing rewards or punishments, etc.), lead to their overexploitation, because even if the optimum is obtained through mutual cooperation there would be an individual incentive to defect. }%
% \es{Se supone que los bienes comunes, si no están acompañados de condiciones especiales (como la comunicación que permita la coordinación, la memoria que permita aplicar premios o castigos, etc), conducen a su sobreexplotación, porque aunque el óptimo se obtenga a través de la mutua cooperación habría una incentivo individual para desertar. }%
\en{However, probabilistic inference shows that cooperative individuals are favored by multilevel selection. }%
\es{Sin embargo, la inferencia probabilística muestra que los individuos cooperadoras se ven favorecidas mediante selección multinivel. }%
\en{While resource-avoidance strategies can invade entirely cooperative groups by natural selection, such behavior at the same time increases their own individual fluctuations, reducing their own long-term growth rate without the need to introduce penalties. }%
\es{Si bien las estrategias que evitan compartir recursos pueden invadir por selección natural grupos enteramente cooperadores, tal comportamiento aumenta sus propias fluctuaciones individuales, reduciendo su propia tasa de crecimiento a largo plazo sin necesidad de introducir castigos. }%
\en{That is, contrary to the belief established since the mid-20th century in economics, we show that the dynamics of common goods cannot be represented by a prisoner's dilemma payoff matrix. }%
\es{Es decir, en contra de la creencia establecida que en economía se tiene desde mediados del siglo 20, mostramos que los dinámicas de bienes comunes no puede representarse mediante una matriz de pagos del dilema del prisionero. }%
\en{Moreover, contrary to the belief that specialization is too complex a feature to produce a benefit in simple aggregations, we show that as soon as cooperation emerges, an advantage in favor of specialist strategies appears even in groups of size 2. }%
\es{Además, en contra de la creencia de que la especialización es una caracterísitica demasiado compleja para que produzca un beneficio en agregaciones simples, mostramos que apenas surge la cooperación, aparece una ventaja a favor de las estrategias especialistas incluso en grupos de tamaño 2. }%
\en{Since the specialist strategies are individually poorly adapted to the environment, an irreversibility of the evolutionary transition is created. }%
\es{Como la estrategias especialistas están individualmente mal adaptadas al ambiente, se crea una irreversbilidad de la transici evolutivas. }%
\en{Finally, although in this causal model there is no evolutionary advantage in favor of the division of labor, its emergence is viable by evolutionary drift. }%
\es{Finalmene, si bien en este modelo causal no hay una ventaja evolutiva a favor de la división del trabajo, su aparición es viable por deriva evolutiva. }%

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probability theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
%
% \en{Even if our causal model does not assume any kind of process to update resources, probabilistic inference is proportional to the resources obtained through a multiplicative process. }%
% \es{Incluso si nuestro modelo causal no presupone ningún tipo de proceso para actualizar los recursos, la inferencia probabilistica es proporcional a los recursos obtenidos a través de un proceso multiplicativo. }%
% %
\en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
\es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }% 
%

@@

\en{In the last third of the history of the Universe, sometime around 4 billion years ago, a simple form of matter organization capable of self-replication appeared on Earth. }%
\es{En el último tercio de la historia del Universo, en algún momento hace aproximadamente 4000 millones de años, apareció en la tierra una forma de organización de la materia capaz de auto-replicarse. }%
%
\en{The growth of these lineages followed multiplicative and noisy processes: sequences of survival and reproduction rates. }%
\es{El crecimiento de estos linajes siguieron procesos multiplicativos y ruidosos: secuencias de probabilidades de supervivencia y reproducción. }%
%
\en{The errors produced during replication diversified the life forms, and the growth rates of the different strategies favored those better adapted to the environment. }%
\es{Los errores producidos durante la replicación diversificaron las formas de vida, y las tasas de crecimiento de las diferentes estrategias favorecieron a aquellas mejor adaptadas al ambiente. }%
%
\en{From that moment until now, life has acquired an extraordinary complexity, both in terms of cooperation and specialization. }%
\es{Desde aquel momento hasta ahora la vida adquirió una extraordinaria complejidad, tanto a nivel de cooperación como de especialización. }%
%
\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.65\textwidth}
    \includegraphics[width=\linewidth]{auxiliar/images/biomass.jpg}
    \end{subfigure}
    \caption{
    \en{Distribution of biomass on Earth estimated by Bar-On et al.~\cite{barOn2018-biomass}. }
	\es{Distribución de la biomasa en la Tierra estimada por Bar-On et al.~\cite{barOn2018-biomass}. }%
    }
    \label{fig:biomass}
\end{figure}
%
\en{The current complexity of life is the consequence of a series of evolutionary transitions in which entities capable of self-replication after the transition become part of higher level cooperative units~\cite{maynardSmith1995-majorTransitions, szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}. }%
\es{La complejidad actual de la vida es consecuencia de una serie de transiciones evolutivas en las que entidades capaces de autoreplicación luego de la transición pasan a formar parte de unidades cooperativas de nivel superior~\cite{maynardSmith1995-majorTransitions, szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}. }%
%
\en{Some of the paradigmatic transitions are: from prokaryotic to eukaryotic cells; from protozoa to animals, plants and fungi (cell differentiation and emergence of multicellular organisms); and from solitary individuals to societies. }%
\es{Algunas de las transiciones paradigmáticas son: de las células procariotas a las eucariotas; de los protozoa a los animales, plantas y hongos (diferenciación celular y emergencia de organiismos multicelulares); y de los individuos solitarios a las sociedades. }%
%
\en{How to explain this permanent tendency of life in favor of cooperative and specialization? }%
\es{¿Cómo se explica esta tendencia permanente de la vida en favor de la cooperación y la especialización? }%
%``The transition must be explained in terms of inmmediate selection advantage to individual replicators'' szathmary1995-evolutionaryTransitions

@@

}
}
\caption{
\en{The agents start with the same initial resources, update them independently according to the equation \ref{eq:estrategia_base}, and finally redistribute it in equal parts. }%
\es{Los agentes comienzan con los mismos recursos iniciales, los actualizan independientemente de acuerdo con la ecuación \ref{eq:estrategia_base}, y finalmente lo redistribuyen en partes iguales. }%
}
\label{fig:protocolo}
\end{figure}
%
\en{In Figure \ref{fig:ergodicity_cooperation} we show the trajectory of an agent in a cooperating group of size 33. }%
\es{En la figura \ref{fig:ergodicity_cooperation} mostramos la trayectoria de un agente en un grupo cooperador de tamaño 33. }%
%
\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/pdf/ergodicity_cooperation.pdf}
    \end{subfigure}
    \caption{
    \en{The resources of an individual in a cooperative group of size 33 (green line) approaches the arithmetic mean (the black line below the green line).
    As a visual reference we show the geometric mean that drives the growth rate of individuals (blue line). }%
    \es{Los recursos de un individuo en un grupo cooperativo de tamaño 33 (recta verde) se aproxima a la media aritmética (la recta negra debajo de la recta verde).
    Como referencia visual mostramos la media geométrica que gobierna la tasa de crecimiento de los individuos (recta azul). }%
    }
    \label{fig:ergodicity_cooperation}
\end{figure}
%
\en{Fully cooperative groups reduce their fluctuations, generating an increase in the growth rate of all their members. }%
\es{Grupos enteramente cooperadoras reducen sus fluctuaciones, generando un aumento en la tasa de crecimiento de todos sus miembros. }%
% 
% \en{Individuals in fully cooperative groups achieve growth rates equivalent to the average of system states, which in non-ergodic systems is always higher than individual growth rates. }%
% \es{Los individuos de los grupos enteramente cooperadoras logran acceder a tasas de crecimiento equivalentes al promedio de estados del sistema, que en los sistemas no-ergódicos es siempre superior que la tasas de crecimiento individual. }%
%
\en{Peters-Amadou~\cite{peters-cooperation2019.03.04} believe that such increase in the growth rate is sufficient argument to demonstrate the evolutionary advantage of cooperation. }%
\es{Peters-Amadou~\cite{peters-cooperation2019.03.04} consideran que el aumento de la tasa de crecimiento es argumento suficiente para demostrar la ventaja evolutiva de la cooperación. }%
%
\en{However, they do not consider the problem of defection, who say ``our cooperators are unable to break the cooperative pact''. }%
\es{Sin embargo ellos no consideran el problema de la deserción, quienes dicen ``our cooperators are unable to break the cooperative pact''. }%
%
\en{This does not seem to be a minor problem, considering the temptation to stop contributing to the common fund while continuing to receive its benefits. }%
\es{No parece ser un problema menor, teniendo en cuenta la tentación de dejar de aportar al fondo común mientras se siguen recibiendo sus beneficios. }%

@@

\en{To demonstrate the evolutionary advantage of cooperation in the presence of defection it is necessary to consider selection at both the individual and group level. }%
\es{Para demostrar la ventaja evolutiva de la cooperción en presencia de desertores es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{Yaari and Solomon~\cite{yaari2010-cooperationEvolution} already suggest a verbal argument for group selection: while there would be a temptation to defect, fully cooperative groups will do better. }%
\es{Yaari y Solomon~\cite{yaari2010-cooperationEvolution} sugieren ya un argumento verbal de la selección de grupo: si bien habría una tentación para desertar, a los grupos enteramente cooperadores le irá mejor. }%
%
\en{However, they do not provide this argument with a formal proof. }%
\es{Sin embargo no acompañan este argumento con una demostración formal. }%
% Parrafo
\en{The co-author of the concept of evolutionary transitions (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models~\cite{czegel2019-bayesianEvolution}, making use of the isomorphism between evolutionary theory and Bayesian inference~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos~\cite{czegel2019-bayesianEvolution}, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%
% Parrafo
\en{In this paper we demonstrate the evolutionary advantage of cooperation in the presence of defection using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En este trabajo demostramos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que represente evolución bajo selección multinivel. }%
%
\en{To the best of our knowledge, our work would be the first to develop a Bayesian hierarchical model to solve an evolution problem under multilevel selection. }%
\es{Hasta donde sabemos, nuestro trabajo sería el primero en desarrollar un modelo jerárquico bayesiano para resolver un problema de evolución bajo selección multinivel. }%
%
% \en{With this model we demonstrate, as Ole Peters intended, that the advantage of cooperation (now in the presence of defection) is a consequence of the most basic and fundamental assumption of evolutionary theory: that evolutionary processes are multiplicative and noisy. }%
% \es{Con este modelo demostramos, como pretendía Ole Peters, de que la ventaja de la cooperación (ahora en presencia de deserción) es consecuencia del supuesto más básico y fundamental de la teoría de la evolución: que los procesos evolutivos son multiplicativos y ruidosos. }%

\paragraph{\en{Specialization}\es{Especialización}}

\en{Peters-Amadou propose multiplicative processes as the main explanation of major evolutionary transitions. }%
\es{Peters-Amadou proponen a los procesos multiplicativos como la principal explicación de la transiciones evolutvas mayores. }%
\en{However, they believe that multiplicative processes can only explain the advantage of cooperation, but not the advantage of specialization. }%
\es{Sin embargo, ellos creen que los procesos multiplicativos sólo pueden explicar la ventaja de la cooperación, pero no la ventaja de la especialización. }%
% \en{Ole Peters believes that specialization is a too complex property to produce a benefit in simple aggregations of, for example, two agents. }
% \es{Ole Peters considera a la especialización como una caracterísitica demasiado compleja para que produzca un beneficio en agregaciones simples de por ejemplo dos agentes. }%
% \en{He says the ``[advantage of cooperation] it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }
% \es{Dice que la ventaja de la cooperación: ``it may explain the transition from single cells to bicellular organisms, too small and simple to benefit from new function or specialization.'' }%
% \en{He believes that other explanations of evolutionary transitions, such as the emergence of specialization, are implausible. }%
% \es{El cree que otras explicaciones de las transiciones evolutivas, como la emergencia de la especialización, son implausibles. }%
% \en{According to Ole Peters, an important evolutionary phenomenon on which his analysis can shed new light is the transition to multicellularity. }%
% \es{Según Ole Peters, un importante fenómeno evolutivo sobre el que su análisis puede arrojar nueva luz es la transición a la multicelularidad. }%
% \en{But, to explain evolutionary transitions, it is necessary to demonstrate the evolutionary advantage of cooperation in the presence of defection, but also the advantage of specialization. }%
% \es{Luego, para explicar las transiciones evolutivas es necesario demostrar la ventaja evolutiva de la cooperación en presencia de deserción, pero también la ventaja de la especiliazación. }%
% \en{However evolutionary transitions not only involve cooperation, it also involves specialization. }%
% \es{Sin embargo, las transiciones evolutivas no involucran sólo cooperación, involucra también especialización. }%
% % Parrafo
\en{In this paper we show that the same minimal assumptions used by Yaari and Peters to explain the evolutionary advantage of cooperation also allow us to explain the evolutionary advantage of specialization, even for the simplest aggregations of two agents. }%
\es{En este trabajo demostramos que los mismos supuestos mínimos que utilizam Yaari y Peters para explicar la ventaja evolutiva de la cooperación, permiten también explicar la ventaja evolutiva de especialización, incluso para las agregaciones más simples de dos agentes. }%
\en{Once cooperation emerges, individually maladapted strategies (specialists) manage to outperform, even organized in small groups, both individually well-adapted strategies (generalists) and their cooperative groups of infinite size. }%
\es{Una vez que la cooperación emerge, las estrategias individualmente mal adaptadas (especialistas) consigen superar, incluso organizadas en grupos pequeños, tanto a las estrategias bien adaptas individualmemte (generalistas) como a sus grupos cooperativos de tamaño infinito. }%
\en{Since the specialist strategies are individually poorly adapted to the environment, an irreversibility of the evolutionary transition is created. }%
\es{Como la estrategias especialistas están individualmente mal adaptadas al ambiente, se crea una irreversbilidad de la transici evolutivas. }%
\en{It is extraordinary that such an essential assumption, i.e. the multiplicative nature of probability and evolutionary theories, has such fundamental conclusions for understanding the emergence of the complexity of life. }%
\es{Es extraordinario que un supuesto tan esencial, i.e. la naturaleza multiplicativa de las teorías probabilisticas y evolutivas, tenga conclusiones tan fudamentales para entender la emergencia de la complejidad de la vida. }%
% \en{Because the evolutionary advantage of cooperation and specialization is a direct consequence of the non-ergodicity of the multiplicative processes to which life is subject, we support Ole Peters' idea of considering it the first cause of major evolutionary transitions. }%
% \es{Debido a que la ventaja evolutiva de la cooperación y la especialización es consecuencia directa de la no-ergodicidad de los procesos multiplicativos a los que está sujeto la vida, apoyamos la idea de Ole Peters de consideradala la primera causa de las transiciones evolutivas mayores. }%

@@

\en{In this section we present the isomorphism between probabilitic and evolutionary theory. }%
\es{En esta sección presentamos el isomorfismo entre las teorías de la probabilidad y de la evolución. }%
%
\en{In the \ref{sec:resuts} section we will poof the evolutionary advantage of cooperation and specialization using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En la sección \ref{sec:results} demostraremos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que representa evolución bajo selección multinivel. }%
%

@@

\en{if their social behavior is cooperative, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (the first $n$ individuals in the region are cooperators and the rest defectors); }%
\es{si su comportamiento social es cooperador, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (los primeros $n$ individuos de la región son cooperadores y el resto desertores); }%

>>>>

%opening
\title{\vspace{-2.2cm}Multilevel selection in causal models: the multiplicative nature of evolutionary and probabilistic selection processes as the general driver for the emergence of cooperation and specialization (Summary paper).\footnote{Add stable link to this summary file} }
\author{Gustavo Landfried$^1$}
\affil{\small 1. Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Computaci\'on. Buenos Aires, Argentina}
\affil{\en{Mail}\es{Correo:} \url{glandfried@dc.uba.ar}}

@@

\begin{abstract}

\en{To explain major evolutionary transitions, it is necessary to demonstrate the advantage of cooperation, specialization and their irreversibility. }%
\es{Para explicar las grandes transiciones evolutivas, es necesario demostrar la ventaja de la cooperación, la especialización y la irreversibilidad de las mismas. }%
%
\en{For this purpose it is necessary to consider selection at both the individual and group level. }%
\es{Para este propósito es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{The co-author of the concept of evolutionary transitions (Szathmáry) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models, making use of the isomorphism between evolutionary theory and Bayesian inference. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmary) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana. }%
%
\en{However, the proposal remains open. }%
\es{Sin embargo, la propuesta sigue abierta. }%
%
% \en{Moreover, the models do not express a causal relationship between the variables. }%
% \es{Además, los modelos no expresan una relación causal entre las variables. }%

@@

\en{In this paper we specify a probabilistic causal model, in which individuals are affected by the environment and by the social behaviors of cooperation and defection of their context. }%
\es{En este trabajo especificamos un modelo causal probabilístico, en el que los individuos se ven afectados por el ambiente y por los comportamientos sociales de cooperación y deserción de su contexto. }%
%
\en{Under this minimal set of hypotheses, where we consider \emph{unconditionally} cooperative individuals who generate a common good that can be exploited by defecting individuals without receiving some kind of punishment in return (e.g. end of cooperation), probabilistic inference shows that cooperative individuals are favored by multilevel selection. }%
\es{Bajo este conjunto mínimo de hipótesis, donde consideramos individuos \emph{incondicionalmente} cooperadores que generan un bien común que puede ser explotado por individuos desertores sin que reciban a cambio algún tipo de castigo (e.g fin de la cooperación), la inferencia probabilística muestra que los individuos cooperadoras se ven favorecidas mediante selección multinivel. }%
%
\en{In addition, we show that as soon as cooperation emerges, an advantage in favor of specialist strategies appears. }%
\es{Además, mostramos que apenas surge la cooperación, aparece una ventaja a favor de las estrategias especialistas. }%
%
\en{Since the specialist strategies are individually poorly adapted to the environment, an irreversibility of the evolutionary transition is created. }%
\es{Como la estrategias especialistas están individualmente mal adaptadas al ambiente, se crea una irreversbilidad de la transici evolutivas. }%

@@

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probability theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
% %
% \en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
% \es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }%

@@

\en{In the last third of the history of the Universe, sometime around 4 billion years ago, a simple form of matter organization capable of self-replication appeared on Earth. }%
\es{En el último tercio de la historia del Universo, en algún momento hace aproximadamente 4000 millones de años, apareció en la tierra una forma de organización de la materia capaz de auto-replicarse. }%
%
\en{The growth of these lineages followed multiplicative and noisy processes: sequences of survival and reproduction rates. }%
\es{El crecimiento de estos linajes siguieron procesos multiplicativos y ruidosos: secuencias de probabilidades de supervivencia y reproducción. }%
%
\en{The errors produced during replication diversified the life forms, and the growth rates of the different strategies favored those better adapted to the environment. }%
\es{Los errores producidos durante la replicación diversificaron las formas de vida, y las tasas de crecimiento de las diferentes estrategias favorecieron a aquellas mejor adaptadas al ambiente. }%
%
\en{From that moment until now, life has acquired an extraordinary complexity, both in terms of cooperation and specialization. }%
\es{Desde aquel momento hasta ahora la vida adquirió una extraordinaria complejidad, tanto a nivel de cooperación como de especialización. }%

\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.65\textwidth}
    \includegraphics[width=\linewidth]{auxiliar/images/biomass.jpg}
    \end{subfigure}
    \caption{
    \en{Distribution of biomass on Earth estimated by Bar-On et al.~\cite{barOn2018-biomass}. }
	\es{Distribución de la biomasa en la Tierra estimada por Bar-On et al.~\cite{barOn2018-biomass}. }%
    }
    \label{fig:biomass}
\end{figure}
%
\en{The current complexity of life is the consequence of a series of evolutionary transitions in which entities capable of self-replication after the transition become part of higher level cooperative units~\cite{maynardSmith1995-majorTransitions, szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}. }%
\es{La complejidad actual de la vida es consecuencia de una serie de transiciones evolutivas en las que entidades capaces de autoreplicación luego de la transición pasan a formar parte de unidades cooperativas de nivel superior~\cite{maynardSmith1995-majorTransitions, szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}. }%
%
\en{Some of the paradigmatic transitions are: from prokaryotic to eukaryotic cells; from protozoa to animals, plants and fungi (cell differentiation and emergence of multicellular organisms); and from solitary individuals to societies. }%
\es{Algunas de las transiciones paradigmáticas son: de las células procariotas a las eucariotas; de los protozoa a los animales, plantas y hongos (diferenciación celular y emergencia de organiismos multicelulares); y de los individuos solitarios a las sociedades. }%
%
\en{How to explain this permanent tendency of life in favor of cooperative and specialization? }%
\es{¿Cómo se explica esta tendencia permanente de la vida en favor de la cooperación y la especialización? }%
%``The transition must be explained in terms of inmmediate selection advantage to individual replicators'' szathmary1995-evolutionaryTransitions

@@

}
}
\caption{
\en{The agents start with the same initial resources, update them independently according to the equation \ref{eq:estrategia_base}, and finally redistribute it in equal parts. }%
\es{Los agentes comienzan con los mismos recursos iniciales, los actualizan independientemente de acuerdo con la ecuación \ref{eq:estrategia_base}, y finalmente lo redistribuyen en partes iguales. }%
}
\label{fig:protocolo}
\end{figure}
%
\en{In Figure \ref{fig:ergodicity_cooperation} we show the trajectory of an agent in a cooperating group of size 33. }%
\es{En la figura \ref{fig:ergodicity_cooperation} mostramos la trayectoria de un agente en un grupo cooperador de tamaño 33. }%
%
\begin{figure}[ht!]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=\linewidth]{figures/pdf/ergodicity_cooperation.pdf}
    \end{subfigure}
    \caption{
    \en{The resources of an individual in a cooperative group of size 33 (green line) approaches the arithmetic mean (the black line below the green line).
    As a visual reference we show the geometric mean that drives the growth rate of individuals (blue line). }%
    \es{Los recursos de un individuo en un grupo cooperativo de tamaño 33 (recta verde) se aproxima a la media aritmética (la recta negra debajo de la recta verde).
    Como referencia visual mostramos la media geométrica que gobierna la tasa de crecimiento de los individuos (recta azul). }%
    }
    \label{fig:ergodicity_cooperation}
\end{figure}
%
\en{Fully cooperative groups reduce their fluctuations, generating an increase in the growth rate of all their members. }%
\es{Grupos enteramente cooperadoras reducen sus fluctuaciones, generando un aumento en la tasa de crecimiento de todos sus miembros. }%
%

@@

\en{To demonstrate the evolutionary advantage of cooperation in the presence of defection it is necessary to consider selection at both the individual and group level. }%
\es{Para demostrar la ventaja evolutiva de la cooperción en presencia de desertores es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{The co-author of the concept of evolutionary transitions (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models~\cite{czegel2019-bayesianEvolution}, making use of the isomorphism between evolutionary theory and Bayesian inference~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmáry~\cite{szathmary1995-evolutionaryTransitions, szathmary2015-evolutionaryTransitions}) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos~\cite{czegel2019-bayesianEvolution}, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana~\cite{harper2009-replicatorAsInference,shalizi2009-replicatorAsInference}. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%
% 
\en{In this paper we demonstrate the evolutionary advantage of cooperation in the presence of defection using a probabilistic causal model representing evolution under multilevel selection. }%
\es{En este trabajo demostramos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo causal probabilistico que represente evolución bajo selección multinivel. }%
%
\en{To the best of our knowledge, our work would be the first to develop a Bayesian hierarchical model to solve an evolution problem under multilevel selection. }%
\es{Hasta donde sabemos, nuestro trabajo sería el primero en desarrollar un modelo jerárquico bayesiano para resolver un problema de evolución bajo selección multinivel. }%
%
% \en{With this model we demonstrate, as Ole Peters intended, that the advantage of cooperation (now in the presence of defection) is a consequence of the most basic and fundamental assumption of evolutionary theory: that evolutionary processes are multiplicative and noisy. }%
% \es{Con este modelo demostramos, como pretendía Ole Peters, de que la ventaja de la cooperación (ahora en presencia de deserción) es consecuencia del supuesto más básico y fundamental de la teoría de la evolución: que los procesos evolutivos son multiplicativos y ruidosos. }%

@@

\en{In this section we present the isomorphism between probabilitic and evolutionary theory. }%
\es{En esta sección presentamos el isomorfismo entre las teorías de la probabilidad y de la evolución. }%
%
\en{In section \nameref{sec:results} we will prove the evolutionary advantage of cooperation, specialization and their irreversibility using a Bayesian hierarchical model with a causal interpretation representing evolution under multilevel selection. }%
\es{En la sección \nameref{sec:results} demostraremos la ventaja evolutiva de la cooperación, la especialización y su irreversbilidad mediante un modelo jerárquico bayesiano con una interpretación causal que representa evolución bajo selección multinivel. }%
%

@@

\en{and if their social behavior is cooperative, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (the first $n$ individuals in the region are cooperators and the rest defectors); }%
\es{y si su comportamiento social es cooperador, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (los primeros $n$ individuos de la región son cooperadores y el resto desertores); }%
